\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-13 paper.qmd
Initial flow of finding the best model for PBMC3k

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     #aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "none", 
     legend.title = element_text(size=5), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm')
   )
)
interior_annotation <- function(label, position = c(0.92, 0.92)) {
  annotation_custom(grid::textGrob(label = label,
      x = unit(position[1], "npc"), y = unit(position[2], "npc"),
      gp = grid::gpar(cex = 1, col="grey70")))
}
```

@@

## Compute error
error_df <- augment(
  df_bin_centroids = df_bin_centroids_pbmc,
  df_bin = df_bin_pbmc,
  training_data = training_data_pbmc,
  newdata = NULL,
  type_NLDR = "PaCMAP",
  col_start = "PC_")

@@

@@

img_sample <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +
  facet_wrap(~ factor(instance, levels = c(3710, 2391, 2385, 5030, 6475,
                         7679, 3568, 1312, 3311, 1097,
                         3552, 7853, 6489, 7689, 6690,
                        1380, 6057, 2347, 5946, 3355,
                        4175, 3997, 5378, 387, 1854,
                        614, 3079, 1762, 5239, 3723,
                        5748, 728, 7419, 7794, 6233,
                        33, 2485, 5998, 318, 1761,
                        5690, 165, 517, 6935, 2682,
                        4962, 2264, 5563, 6369, 559,
                        5188, 4849, 2666, 3448, 3055,
                        3120, 6869, 6345, 4470, 7147)), ncol = 5) +
  scale_fill_continuous_sequential(palette = "Grays") +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank())
```

@@

imge_error_sample_within <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ factor(instance, levels = c( 4699, 4785, 5173, 6671, 2186, 4636, 
                                            4882, 773, 2825, 6169, 6948, 5987, 
                                            1282, 7715, 622, 2989, 6912, 4370, 
                                            1227, 4042, 2444, 836, 1654)), 
             ncol = 5) +
  scale_fill_continuous_sequential(palette = "Grays") +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank())
```

@@

imge_error_sample <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 2) +
  scale_fill_continuous_sequential(palette = "Grays") +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank())
```

>>>>

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     #aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "bottom", 
     legend.title = element_blank(), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm')
   )
)
interior_annotation <- function(label, position = c(0.92, 0.92)) {
  annotation_custom(grid::textGrob(label = label,
      x = unit(position[1], "npc"), y = unit(position[2], "npc"),
      gp = grid::gpar(cex = 1, col="grey70")))
}
```

@@

## Compute error
error_df <- augment(
  df_bin_centroids = df_bin_centroids_pbmc,
  df_bin = df_bin_pbmc,
  training_data = training_data_pbmc,
  newdata = NULL,
  type_NLDR = "UMAP",
  col_start = "PC_")

@@

<!--compute absolute error for different parameter choices-->
```{r}
error_df_umap <- read_csv("data/pbmc3k/error_df_umap.csv", col_names = FALSE)
names(error_df_umap) <- c("error", "mse", "bin1", "bin2", "b", "b_non_empty", 
                     "method", "n_neighbors", "min_dist")

error_df_umap <- error_df_umap |>
  mutate(type = paste0("n_neighbors: ", n_neighbors, ", min_dist: ", min_dist)) |>
  select(-c(n_neighbors, min_dist))

error_df_tsne <- read_csv("data/pbmc3k/error_df_tsne.csv", col_names = FALSE)
names(error_df_tsne) <- c("error", "mse", "bin1", "bin2", "b", "b_non_empty", 
                          "method", "perplexity")

error_df_tsne <- error_df_tsne |>
  mutate(type = paste0("perplexity: ", perplexity)) |>
  select(-perplexity)

error_df <- bind_rows(error_df_umap, error_df_tsne)

error_plot_pbmc <- ggplot(error_df, 
                              aes(x = b_non_empty, 
                                  y = log(error), 
                                  group = type, 
                                  colour = type)) + 
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = 137, linetype="solid", 
             color = "black", size=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99',
                              '#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a', 
                              '#b15928', '#ffff99')) +
  ylab("log(absolute error)") +
  xlab("number of non-empty bins")

```{r}
#| fig-cap: "Absolute error from UMAP and tSNE applied to training PBMC3k dataset with diffferent parameter choices. What is the best parameter choice to create the model? The residual plot have a steep slope at the beginning, indicating that a smaller number of non-empty bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of non-empty bins generates a smaller error. Using the elbow method, it was observed that when the number of non-empty bins is set to 137, the lowest error occurred with the parameters perplexity: 30."
#| label: fig-abserrorPBMC
#| fig-pos: H
#| out-width: 100%

error_plot_pbmc
```

<!--best choice-->

```{r}
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_30.rds")
pbmc_scaled_obj <- gen_scaled_data(
  data = tsne_pbmc)
tsne_pbmc_scaled <- pbmc_scaled_obj$scaled_nldr

tsne_pbmc <- tsne_pbmc_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2)) +
  geom_point(alpha=0.5)
```

<!--Fit the best model and compute error-->
```{r}
## Compute hexbin parameters
num_bins_x_pbmc <- 15
lim1 <- pbmc_scaled_obj$lim1
lim2 <- pbmc_scaled_obj$lim2
r2_pbmc <- diff(lim2)/diff(lim1)

pbmc_model <- fit_highd_model(
  training_data = training_data_pbmc,
  emb_df = tsne_pbmc_scaled,
  bin1 = num_bins_x_pbmc,
  q = 0.1,
  r2 = r2_pbmc,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "PC_"
)

benchmark_pbmc <- 0.1

## Compute error
error_df <- augment(
  df_bin_centroids = df_bin_centroids_pbmc,
  df_bin = df_bin_pbmc,
  training_data = training_data_pbmc,
  newdata = NULL,
  type_NLDR = "tSNE",
  col_start = "PC_")

error_df <- error_df |>
  mutate(type = case_when(
    row_wise_abs_error <= 5 ~ "error 0-5",
    row_wise_abs_error <= 10 ~ "error 5-10",
    row_wise_abs_error <= 15 ~ "error 10-15",
    row_wise_abs_error <= 20 ~ "error 15-20",
    row_wise_abs_error <= 25 ~ "error 20-25",
    .default = "error greter than 25"
  )) |>
  mutate(type = factor(type, levels = c(
    "error 0-5", "error 5-10", "error 10-15", "error 15-20", "error 20-25", 
    "error greter than 25")))

## To join embedding
error_df <- error_df |>
  bind_cols(tsne_pbmc_scaled |>
              select(-ID))

error_plot_pbmc <- error_df |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             color = type,
             group = ID)) +
  geom_point(alpha=0.5,
             size = 0.1) +
  interior_annotation("b", sc_ltr_pos)

<!--bin1 = 15, bin2 = 20, b = 300, non_empty = 137-->
```{r}
#| echo: false
#| fig-cap: "(a) Model generated in the 2D space overlaid on tSNE data, and (b) high-D model error in model space."
#| label: fig-model-pbmc
#| fig-pos: H

@@

img_sample <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +
  facet_wrap(~ factor(instance, levels = c(3710, 2391, 2385, 5030, 6475,
                         7679, 3568, 1312, 3311, 1097,
                         3552, 7853, 6489, 7689, 6690,
                        1380, 6057, 2347, 5946, 3355,
                        4175, 3997, 5378, 387, 1854,
                        614, 3079, 1762, 5239, 3723,
                        5748, 728, 7419, 7794, 6233,
                        33, 2485, 5998, 318, 1761,
                        5690, 165, 517, 6935, 2682,
                        4962, 2264, 5563, 6369, 559,
                        5188, 4849, 2666, 3448, 3055,
                        3120, 6869, 6345, 4470, 7147)), ncol = 5) +
  scale_fill_continuous_sequential(palette = "Grays") +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank(), 
        legend.position = "none")
```

@@

imge_error_sample_within <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ factor(instance, levels = c( 4699, 4785, 5173, 6671, 2186, 4636, 
                                            4882, 773, 2825, 6169, 6948, 5987, 
                                            1282, 7715, 622, 2989, 6912, 4370, 
                                            1227, 4042, 2444, 836, 1654)), 
             ncol = 5) +
  scale_fill_continuous_sequential(palette = "Grays") +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank(),
        legend.position = "none")
```

@@

imge_error_sample <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +  
  facet_wrap(~ instance, ncol = 2) +
  scale_fill_continuous_sequential(palette = "Grays") +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank(),
        legend.position = "none")
```

### 2024-05-13 paper.tex
Initial flow of finding the best model for PBMC3k

>>>>

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-abserrorPBMC-1.pdf}

\caption{\label{fig-abserrorPBMC}Absolute error from UMAP and tSNE
applied to training PBMC3k dataset with diffferent parameter choices.
What is the best parameter choice to create the model? The residual plot
have a steep slope at the beginning, indicating that a smaller number of
non-empty bins causes a larger amount of error. Then, the slope
gradually declines or level off, indicating that a higher number of
non-empty bins generates a smaller error. Using the elbow method, it was
observed that when the number of non-empty bins is set to 137, the
lowest error occurred with the parameters perplexity: 30.}

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-model-pbmc-1.pdf}

\caption{\label{fig-model-pbmc}(a) Model generated in the 2D space
overlaid on tSNE data, and (b) high-D model error in model space.}
